---
title: Constructing control regions for enrichment analysis
author: Jean Monlong
date: March 3, 2016
output:
   md_document:
     fig_width: 10
     fig_height: 5
     variant: markdown_github
---

# Constructing control regions for enrichment analysis

## Introduction

When investigating enrichment in specific genomic features, it is common to compare the regions of interest with some controls regions. **A simple approach** to construct control regions could be to **randomly select regions across the genome**. Additionally, it is important for the control regions to have the **same size distribution**.

However, a random distribution across the genome is usually **not realistic**. Likely, regions in the genome were not tested because inaccessible or not included in the analysis. Moreover, **you might want to control for some patterns and look for more**.

For example, we first observed enrichment of CNVs in low-mappability regions. We then wanted to test additional enrichment in different repeat classes. Because repeats are enriched in low-mappability regions, repeats will likely be seen enriched in CNVs. We want to avoid spurious correlation and control for the low-mappability enrichment. By constructing control regions with the same low-mappability enrichment we can can test additional enrichment in the different repeat classes without being biased by the relation between low-mappability regions and repeats.


## Constructing control regions with *PopSV* package

First we load the package and retrieve some annotations to play with.

```{r, message=FALSE, warning=FALSE}
library(PopSV)
library(AnnotationHub)
ah = AnnotationHub()
genes = ah[["AH49010"]] ## Genes
dgv = ah[["AH5120"]] ## SVs from DGV
dgv = dgv[sample.int(length(dgv), 1e4)] ## Reduce to 10K random SVs
```

We imported a gene annotation and 10 thousands SVs from DGV. If we want to construct control regions that fit the SV size and overlap with genes, we run:

```{r, message=FALSE}
dgv.cont = draw.controls(dgv, list(gene=genes), chr.prefix="chr")
```

Now let's verify that the size distribution is the same. By construction it should be **exactly** the same.

```{r, message=FALSE}
library(ggplot2)
size.df = rbind(data.frame(reg="dgv", size=width(dgv)),
                 data.frame(reg="control", size=width(dgv.cont)))
ggplot(size.df, aes(x=size, fill=reg)) + geom_histogram(position="dodge")
ggplot(size.df, aes(x=size, fill=reg)) + geom_histogram(position="dodge") + scale_x_log10()
```

And that the input and output regions **overlap** genes **similarly**. By default an **approximation** is used and might lead to small differences. To switch off the approximation see further down.

```{r, message=FALSE}
mean(overlapsAny(dgv, genes))
mean(overlapsAny(dgv.cont, genes))
```

`draw.controls` functions can **accept any number of genomic features to control**. Let's import two additional genomic annotation that we would like to control for our enrichment analysis: assembly gaps and segmental duplications.

```{r, message=FALSE}
gap = ah[["AH6444"]]
segdups = ah[["AH5121"]]
dgv.cont2 = draw.controls(dgv, list(gene=genes, gap=gap, sd=segdups), chr.prefix="chr")
```

Again, the size distribution must be the same:

```{r, message=FALSE}
## Same size distribution ?
size.df = rbind(data.frame(reg="dgv", size=width(dgv)),
                 data.frame(reg="control", size=width(dgv.cont2)))
ggplot(size.df, aes(x=size, fill=reg)) + geom_histogram(position="dodge")
ggplot(size.df, aes(x=size, fill=reg)) + geom_histogram(position="dodge") + scale_x_log10()
```

And the overlap with the three different genomic annotations similar.

```{r, message=FALSE}
## Same overlap with features ?
mean(overlapsAny(dgv, genes))
mean(overlapsAny(dgv.cont2, genes))
mean(overlapsAny(dgv, gap))
mean(overlapsAny(dgv.cont2, gap))
mean(overlapsAny(dgv, segdups))
mean(overlapsAny(dgv.cont2, segdups))
```

If we had used the first set of control regions (only genes overlap control) the gap and segmental duplication overlap proportions wouldn't match.

```{r, message=FALSE}
mean(overlapsAny(dgv.cont, segdups))
mean(overlapsAny(dgv.cont, gap))
```

## Controlling for the distance to a feature

In addition to controlling for the overlap to a set of feature, `draw.controls` can also **control for the distance to one feature**. Although we can control for overlap to several feature the distance control is more complex to multiplex and for now we can only control for distance to one feature. 

```{r, message=FALSE}
dgv.cont3 = draw.controls(dgv, list(gene=genes, sd=segdups), chr.prefix="chr", dist.gr=gap)
```

Again, the size distribution must be the same:

```{r, message=FALSE}
## Same size distribution ?
size.df = rbind(data.frame(reg="dgv", size=width(dgv)),
                 data.frame(reg="control", size=width(dgv.cont3)))
ggplot(size.df, aes(x=size, fill=reg)) + geom_histogram(position="dodge")
ggplot(size.df, aes(x=size, fill=reg)) + geom_histogram(position="dodge") + scale_x_log10()
```

And the overlap with the three different genomic annotations similar.

```{r, message=FALSE}
## Same overlap with features ?
mean(overlapsAny(dgv, genes))
mean(overlapsAny(dgv.cont3, genes))
mean(overlapsAny(dgv, segdups))
mean(overlapsAny(dgv.cont3, segdups))
```

Finally let's check that we now control for the *distance* to a gap.


```{r, message=FALSE}
## Same distance to gap ?
dist.df = rbind(data.frame(reg="dgv", dist=as.data.frame(distanceToNearest(dgv, gap))$distance),
    data.frame(reg="simple control", dist=as.data.frame(distanceToNearest(dgv.cont2, gap))$distance),
    data.frame(reg="distance control", dist=as.data.frame(distanceToNearest(dgv.cont3, gap))$distance))
ggplot(dist.df, aes(x=dist, fill=reg)) + geom_histogram(position="dodge") + xlab("distance to a gap")
ggplot(dist.df, aes(x=dist, colour=reg)) + stat_ecdf() + ylab("cumulative proportion of regions") + xlab("distance to a gap")
```

## Approximation or exact overlap ?

In order to speed up the process, the input regions are grouped by size. Control regions are then built  for each group using the average size. **This approximation leads to small differences between feature overlap with input and output regions**. It's **not a problem usually**, especially when the input regions have specific size patterns (e.g. CNVs from a binned genome).

If you want **to force the control regions to have exactly the same overlap with the feature, use** `nb.class=Inf`. However it will **take more time to compute**, especially if the size distribution of the input regions is very diverse. Of note the computation time also increases with additional features to control.

```{r, message=FALSE}
system.time((dgv.cont.approx = draw.controls(dgv, list(gene=genes), chr.prefix="chr")))
system.time((dgv.cont.exact = draw.controls(dgv, list(gene=genes), chr.prefix="chr", nb.class=Inf)))
mean(overlapsAny(dgv, genes))
mean(overlapsAny(dgv.cont.approx, genes))
mean(overlapsAny(dgv.cont.exact, genes))
```


## R session

```{r}
sessionInfo()
```
